
#======================================================================================================
Causal relationships among ecology, subsistence strategies, and social organization in prehistoric farming societies: A Bayesian network approach

Rosy Oh and Man-Suk Oh

Apr 26, 2026

#======================================================================================================


#Install and load necessary packages
```{r}
# ── Bioconductor packages (not available on CRAN) ──────────────────────────────
bioc_req <- c("graph", "RBGL", "Rgraphviz")
bioc_missing <- setdiff(bioc_req, rownames(installed.packages()))
if (length(bioc_missing) > 0) {
  if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  BiocManager::install(bioc_missing)
}

# ── CRAN packages ───────────────────────────────────────────────────────────
cran_req <- c(
  "readxl",   # read data
  "pcalg",    # PC algorithm-based causal discovery
  "igraph",   # graph_from_adjacency_matrix
  "bnlearn",  # Bayesian network learning
  "micd",     # mixCItest: conditional independence test for mixed data
  "tpc",      # tpc() algorithm
  "nnet",     # multinom(): multinomial logistic regression
  "dplyr",
  "tidyverse",
  "GGally",
  "httr"
)
cran_missing <- setdiff(cran_req, rownames(installed.packages()))
if (length(cran_missing) > 0)
  install.packages(cran_missing, repos = "https://cloud.r-project.org")

# ── Load packages ───────────────────────────────────────────────────────────
suppressPackageStartupMessages({
  library(readxl)
  library(pcalg)
  library(igraph)
  library(Rgraphviz)
  library(bnlearn)
  library(micd)
  library(tpc)
  library(nnet)
  library(dplyr)
  library(tidyverse)
  library(GGally)
  library(httr)
})
```



# read data and pre-procssing 
```{r}
# ──read original data ─────────────────────────────────────────────────────────────
# Download file from PLOS ONE (Palacios et al. 2022))
url <- "https://doi.org/10.1371/journal.pone.0276088.s006"
tmp <- tempfile(fileext = ".xlsx")
GET(url, write_disk(tmp, overwrite = TRUE), follow_redirects = TRUE)

# Read with skip=1 to bypass the merged category header row
dat.raw0 <- read_excel(tmp,skip = 1)

# Fix the first two unnamed columns
colnames(dat.raw0)[1:2] <- c("Society", "Region")

# ──Remove Lepcha (erroneous 'A' code in Animal husbandry) ──────────────────────────
dat.raw <- dat.raw0[dat.raw0$Society!="Lepcha", ]

# ──rename variables  ────────────────────────────────────────────────────
names <- c(  "Society", "Region",  "Landscape",  "DistanceCoast", "Elevation", "Slope",
  "AnnualTemp", "cv.AnnualTemp",  "MonthlyPrecip", "cv.MonthlyPrecip",  "MonthlyPrimaryprod", "cv.MonthlyPrimaryprod",  "Hunting", "Gathering", "AnimalHusbandry", "Fishing", "Agriculture",
  "CommSize", "SettlementType", "CommOrg", "HouseholdOrg",  "None",
  "ResourceDiversification", "CropSpecialisation", "ForagingIntensification",
  "Storage", "Transhumance", "TemporalMigration", "PermanentMigration",
  "ExchangeOutSettlement", "ExchangeInSettlement", "Reciprocity"
)

colnames(dat.raw)= names


#─── remove  "None", cv variables, Society, Soc Decision variables───────────────────────
dat.raw <- dat.raw %>%
  select(
    -any_of("None"),                          # remove "None" variable
    -starts_with("cv."),                      # remove variables starting with "cv."
    -all_of(names(dat.raw)[c(1:3,22:ncol(dat.raw))]) # remove Society, Soc Decision variables
  )

dim(dat.raw)
colnames(dat.raw)

#define Ecology, Subsistence, SocialOrg variables
Ecology=names(dat.raw)[1:6]
Subsistence=names(dat.raw)[7:11]
SocialOrg=names(dat.raw)[12:15]


#──── convert Ecology as numeric ─────────────────────────────────────────────────────────
dat.raw[, Ecology] <- lapply(dat.raw[, Ecology], 
                              function(x) suppressWarnings(as.numeric(x))) 
summary(dat.raw)

#───── convert Subsistence (interval) to score (numeric)───────────────────
dat.score=dat.raw

map <- c(  "0-5" = 2.5,  "6-15" = 10.5,  "16-25" = 20.5,"26-35" = 30.5,  "36-45" = 40.5,
  "46-55" = 50.5,  "56-65" = 60.5,  "66-75" = 70.5,  "76-85" = 80.5,  "86-100" = 93)

dat.score[,  Subsistence] <- lapply(dat.score[,  Subsistence], function(x) {
  x <- trimws(as.character(x))          # handle potential factor/whitespace
  out <- unname(map[x])                # mapping (NA if no match)
  as.numeric(out)
})


# SocialOrg
#────── convert CommSize(interval)  to score ──────────────────────────────────────
map_cs <- c(
  "<50"      = 25,
  "50-99"    = 75,
  "100-199"  = 150,
  "200-399"  = 300,
  ">400"     = 450
)

dat.score$CommSize <- trimws(as.character(dat.score$CommSize))
dat.score$CommSize <- unname(map_cs[dat.score$CommSize])
dat.score$CommSize <- as.numeric(dat.score$CommSize)

#───────── factorize categorical SocialOrg valriables ────────────────────────────────────
sub_SocialOrg <- setdiff(SocialOrg, "CommSize")
dat.score[, sub_SocialOrg] <- lapply(dat.score[, sub_SocialOrg], function(x) {
  x <- na_if(trimws(as.character(x)), "NA")  # string "NA" → true NA
  as.factor(x)
})

#───────── remove obs with missing values ──────────────────────────────────────────────────────
id_na <- which(rowSums(is.na(dat.score)) > 0)
dat.score <- dat.score %>% drop_na() %>% data.frame()
dim(dat.score) 

#write.csv(dat.score, file="data/dat_score.csv", row.names = F)
#dat.score=read.csv("data/dat_score.csv")

```


#Normalize and standardize (mean=0, sd=1) continuous variables 
```{r}
#Normalize
source("fn_normalize.r")
cols.conti <- which(sapply(dat.score, is.numeric))
res <- normalize(dat.score, cols.conti)
#res$summary     
dat.trans=res$dat_trans   # transformed data

# standardize
dat.trans[, cols.conti] <- scale(dat.trans[, cols.conti])

#cor_mat <- cor(dat.trans[, cols.conti], use = "pairwise.complete.obs", method = "pearson")

```



# blacklist
```{r}

# prepare 
forb_all <- matrix(FALSE, nrow=ncol(dat.trans), ncol=ncol(dat.trans))
rownames(forb_all) <- colnames(forb_all) <- names(dat.trans)

# ──main structure───────────────────────────────────────────────────────
# Allow Ecology -> Subsistence -> SocialOrg,  not the reverse

  forb_all[Subsistence, Ecology] <- TRUE
  forb_all[SocialOrg, Ecology] <- TRUE
  forb_all[SocialOrg, Subsistence] <- TRUE
  
# ──within Ecology───────────────────────────────────────────────────────  
# Geography -> Topography -> Climate -> Productivity, not the reverse 
  
  Geography <- c("DistanceCoast", "Elevation")
  Topography <- c("Slope")
  Climate <- c("AnnualTemp",  "MonthlyPrecip"  )
  Productivity <- c("MonthlyPrimaryprod" )

  forb_all["DistanceCoast", "Elevation"] <- TRUE
  forb_all["Elevation", "DistanceCoast"] <- TRUE
  forb_all[Topography, Geography] <- TRUE
  forb_all[Climate, Geography] <- TRUE
  forb_all[Climate, Topography] <- TRUE
  forb_all[Productivity, Geography] <- TRUE
  forb_all[Productivity, Topography] <- TRUE
  forb_all[Productivity, Climate] <- TRUE
  
# ──within Subsistence───────────────────────────────────────────────────────  
# Agriculture -> AnimalHusbandry -> {Hunting",  "Fishing} -> Gathering, not the reverse 
  
   forb_all["AnimalHusbandry", "Agriculture"] <- TRUE
   forb_all[c("Hunting",  "Fishing",  "Gathering" ),c( "Agriculture", "AnimalHusbandry")] <- TRUE
   forb_all["Gathering", c("Hunting", "Fishing" )]    <- TRUE     

# Select only analysis variables (Ecology + Subsistence + SocialOrg)
# and subset blacklist matrix to match dimensions and order of dat exactly
 cols <- c(Ecology, Subsistence, SocialOrg)
 dat  <- dat.trans %>% select(all_of(cols))
 forb <- forb_all[cols, cols]

```


#Leran Bayesian Network Structure using bagging
```{r}

# Fit a single BN to one bootstrap sample using the tpc() algorithm
# (a parallel, stable version of the PC algorithm for mixed data)
fit_one <- function(dat_sub, forb, alpha_level, numCores) {
  fit <- tpc(
    suffStat    = dat_sub,
    indepTest   = mixCItest,      # mixed CI test for continuous + categorical
    labels      = names(dat_sub),
    alpha       = alpha_level,    # significance level for conditional independence tests
    forbEdges   = forb,           # blacklist matrix encoding domain constraints
    maj.rule    = TRUE,           # use majority rule for edge orientation
    skel.method = "stable.parallel",
    numCores    = 4
  )
  return(fit)
}

# ── Bootstrap settings ───────────────────────────────────────────────────────
seed=4026
dat <- dat.trans
B   <- 50                       # number of bootstrap replicates
p   <- ncol(dat)
vn  <- names(dat)

# Storage: only successfully learned DAGs (cycle-free networks) are kept
boot_nets        <- list()
successful_boots <- 0
skipped_boots    <- c()           # indices of replicates that produced cycles

# ── Main bootstrap loop ──────────────────────────────────────────────────────
for (b in 1:B) {
  # Draw 80% of observations with replacement
  idx <- sample.int(nrow(dat), size = floor(0.8 * nrow(dat)), replace = TRUE)
  fit <- fit_one(dat[idx, ], forb = forb, alpha_level = 0.1, numCores = numCores)

  # Convert to bn object; skip if tpc() produced a cyclic graph
  tryCatch({
    g                <- as.bn(fit, check.cycles = TRUE)
    successful_boots <- successful_boots + 1
    boot_nets[[successful_boots]] <- g
  }, error = function(e) {
    cat("Bootstrap", b, "skipped: contains cycles\n")
    skipped_boots <<- c(skipped_boots, b)  
  })
}
cat(sprintf("Completed: %d/%d replicates accepted (%d skipped)\n",
            successful_boots, B, length(skipped_boots)))

# ── Arc strength calculation ─────────────────────────────────────────────────
source("fn_calculate_arc_strength.r")
arc_strength <- calculate_arc_strength(boot_nets, vn)

summary(arc_strength$strength) #strength and direction should both be in [0, 1]

# ── Arc selection ────────────────────────────────────────────────────────────
# Retain arcs appearing in >= 30% of replicates with consistent direction (>= 50%)
threshold     <- 0.3
selected_arcs <- arc_strength[
  arc_strength$strength  >= threshold &
  arc_strength$direction >= 0.5, ]
print(selected_arcs)

# write.csv(arc_strength, "data/arc_strength_direction.csv", row.names = FALSE)
```
# Construct consensus DAG from selected arcs 
```{r}
# Initialise an empty graph on all nodes
avg_graph <- empty.graph(vn)

# Add selected arcs one by one, skipping any that would create a cycle
for (k in 1:nrow(selected_arcs)) {
  test_graph <- set.arc(avg_graph,
                        from         = as.character(selected_arcs$from[k]),
                        to           = as.character(selected_arcs$to[k]),
                        check.cycles = TRUE)
  
  if (acyclic(test_graph, directed = TRUE)) {
    avg_graph <- test_graph
  } else {
    message("Skipped arc: ", selected_arcs$from[k], " -> ",
            selected_arcs$to[k], " (would create cycle)")
  }
}

# Remove isolated nodes (retain only nodes involved in at least one arc)
connected_nodes <- unique(c(arcs(avg_graph)[, "from"], arcs(avg_graph)[, "to"]))
avg_graph <- subgraph(avg_graph, connected_nodes)
cat("Number of arcs in consensus DAG:", narcs(avg_graph), "\n")

# Save / reload consensus DAG (uncomment reload line as needed)
saveRDS(avg_graph, file = "avg_graph.rds", compress = TRUE)
# avg_graph <- readRDS("avg_graph.rds")

# ── Plot consensus DAG ───────────────────────────────────────────────────────

gR <- graphviz.plot(avg_graph, shape = "rectangle")

# ── Node appearance ──────────────────────────────────────────────────────────
node.attrs <- nodeRenderInfo(gR)
arc.attrs  <- edgeRenderInfo(gR)

# Global node style
node.attrs$col     <- "black"
node.attrs$textCol <- "black"
node.attrs$fontsize <- 14

# Node fill colours by variable group
node.attrs$fill[c(Geography, Topography)] <- "lightgray"    # ecological geography
node.attrs$fill[Climate]                  <- "lightyellow"  # climate
node.attrs$fill[Productivity]             <- "lightyellow"  # primary productivity
node.attrs$fill[Subsistence]              <- "lightgreen"   # subsistence strategies
node.attrs$fill[SocialOrg]               <- "lightblue"    # social organisation

nodeRenderInfo(gR) <- node.attrs

# ── Arc appearance ───────────────────────────────────────────────────────────
# Arcs with direction < 0.60: dashed (weak directional consensus)
# Arcs with direction >= 0.60: solid  (strong directional consensus)
all_arc_names <- names(arc.attrs$lty)

for (k in 1:nrow(selected_arcs)) {
  arc_name <- paste0(selected_arcs$from[k], "~", selected_arcs$to[k])
  
  if (arc_name %in% all_arc_names) {
    arc.attrs$col[arc_name] <- "black"
    arc.attrs$lty[arc_name] <- ifelse(selected_arcs$direction[k] < 0.60,
                                      "dashed", "solid")
  }
}

edgeRenderInfo(gR) <- arc.attrs
renderGraph(gR)

# ── Save plots ───────────────────────────────────────────────────────────────
# File names encode the strength threshold (e.g., threshold=0.3 -> "BN_thr3")
filename <- paste0("figures/BN_thr", as.integer(threshold * 10))

png(paste0(filename, ".png"), res = 300, width = 15, height = 10,
    units = "cm", bg = "transparent")
renderGraph(gR)
dev.off()

pdf(paste0(filename, ".pdf"))
renderGraph(gR)
dev.off()
```



#  Model fitting: estimate local distributions for each node 
```{r}

# Mixed-variable BN: continuous nodes use Gaussian regression,
# discrete nodes use multinomial logistic regression (via nnet::multinom).
# The resulting fitted_list stores one fitted model per node.

fitted_list <- list()

# Identify discrete and continuous nodes present in the consensus DAG
discrete_vars   <- intersect(c("SettlementType", "CommOrg", "HouseholdOrg"),
                             nodes(avg_graph))
continuous_nodes <- setdiff(nodes(avg_graph), discrete_vars)

# ── Continuous nodes: Gaussian conditional distributions ────────────────────
for (node in continuous_nodes) {
  parents <- parents(avg_graph, node)
  
  if (length(parents) == 0) {
    # Root node (no parents): store unconditional mean and SD
    fitted_list[[node]] <- list(
      coef = mean(dat[[node]]),
      sd   = sd(dat[[node]])
    )
  } else {
    # Non-root node: fit linear regression on parent nodes
    formula_str        <- paste(node, "~", paste(parents, collapse = " + "))
    fitted_list[[node]] <- lm(as.formula(formula_str), data = dat)
  }
}

# ── Discrete nodes: multinomial logistic regression ──────────────────────────
# nnet::multinom() handles both purely discrete and mixed (continuous) parents
library(nnet)

for (node in discrete_vars) {
  parents <- parents(avg_graph, node)
  
  if (length(parents) == 0) {
    # Root node: store empirical marginal distribution
    fitted_list[[node]] <- table(dat[[node]]) / nrow(dat)
  } else {
    # Non-root node: fit multinomial logistic regression on parent nodes
    formula_str        <- paste(node, "~", paste(parents, collapse = " + "))
    fitted_list[[node]] <- multinom(as.formula(formula_str),
                                    data  = dat,
                                    trace = FALSE)   # suppress iteration output
  }
}
```



#  Interventional analysis (do-calculus via Monte Carlo simulation)
```{r}
# ── source necessary functions ────────────────────────────────────────────────────

source("fn_simulate_all_Y_interventions.r")
#Functions defined in this file: 
#   simulate_all_Y_interventions() : generate simulations for multiple Y interventions (all at once)
#   simulate_doY_region() : Simulate the joint distribution of all nodes under a specified intervention 
#   draw_Y_empirical(): Sample n values of Y from the observed data under a specified intervention


source("fn_interventional_analysis_ci.r")
#Functions defined in this file: 
#   compute_interventional_analysis_ci(): main wrapper ; computes interventional prob for all X-Y pairs
#   compute_interventional_prob_with_ci() : P(X | do(Y)) and 95% CI when Y is continuous
#   compute_interventional_prob_with_ci_categorical_Y(): P(X | do(Y = k)) and 95% CI when Y is categorical
#   create_results_table()    : organises results into two data frames (continuous Y / categorical Y)
#   plot_ratio(): Plot interventional probability ratios as grouped bar charts 


# ── Define intervention variables (X) and ancestor variables (Y) ─────────────

Subsistence <- c("Hunting", "Gathering", "AnimalHusbandry", "Fishing", "Agriculture")
SocialOrg   <- c("CommSize", "SettlementType", "CommOrg", "HouseholdOrg")

# Retain only X variables that are present in the consensus DAG
X_vars <- intersect(c(Subsistence, SocialOrg), nodes(avg_graph))

# Y variables: all ancestors of X variables in the DAG (i.e., upstream causes)
all_ancestors <- unique(unlist(lapply(X_vars, function(x) {
  bnlearn::ancestors(avg_graph, x)
})))

# exclude X variables that are included in all_ancestors;
all_ancestors <- setdiff(all_ancestors, X_vars)

Y_vars <- all_ancestors


# ── Simulate interventional dist of  X | do(Y = y) ──────────────────────────

## For each Y: intervene at above/below median (continuous) or at each factor level (discrete)
ay_list <- setNames(
  lapply(all_ancestors, function(y) {
    if (is.numeric(dat[[y]])) {
      median(dat[[y]], na.rm = TRUE)   # continuous Y: do(Y > median) vs do(Y <= median)
    } else {
      levels(dat[[y]])                 # # discrete Y: do(Y = k) for each category k
    }
  }),
  all_ancestors
)



sim_results <- simulate_all_Y_interventions(
  fitted_list = fitted_list,
  avg_graph   = avg_graph,
  dat         = dat,
  Y_vars      = all_ancestors,
  ay_list     = ay_list,
  n           = 1e5                    # number of Monte Carlo samples
)

# saveRDS(sim_results, file = "sim_results.rds", compress = TRUE)
# sim_results <- readRDS("sim_results.rds")


# ── Compute interventional ratios and confidence intervals ────────────────────────────

# For each X: intervene at above/below median (continuous) or all factor levels (discrete)
ax_list <- setNames(
  lapply(X_vars, function(x) {
    if (is.numeric(dat[[x]])) {
      median(dat[[x]], na.rm = TRUE)
    } else {
      levels(dat[[x]])
    }
  }),
  X_vars
)


results <- compute_interventional_analysis_ci(
  sim_results = sim_results,
  avg_graph   = avg_graph,
  dat         = dat,
  X_vars      = X_vars,
  ax_list     = ax_list     # if NULL, median of each X is used by default
)

results_tables <- create_results_table(results, digits = 3)

# ── Save interventional results ───────────────────────────────────────────────
if (!is.null(results_tables$continuous_Y)) {
  write.csv(results_tables$continuous_Y,
            "data/interventional_results_continuous_Y.csv", row.names = FALSE)
}
if (!is.null(results_tables$categorical_Y)) {
  write.csv(results_tables$categorical_Y,
            "data/interventional_results_categorical_Y.csv", row.names = FALSE)
}

# ── Plot interventional probability ratios ───────────────────────────────────

# Label each arc as increasing or decreasing effect based on ratio > 1
results_tables$continuous_Y$ratio_ind <- ifelse(
  results_tables$continuous_Y$ratio > 1, "Increase", "Decrease"
)

# Retain only Subsistence and SocialOrg variables present in the results,
# preserving the desired display order for the plot
desired_order          <- c("Agriculture", "Fishing", "AnimalHusbandry", "Hunting", "Gathering")
Subsistence_in_network <- desired_order[desired_order %in%
                            intersect(results_tables$continuous_Y$X, Subsistence)]
SocialOrg_in_network   <- intersect(results_tables$continuous_Y$X, SocialOrg)

plot_ratio(results_tables$continuous_Y, Subsistence_in_network, SocialOrg_in_network)

```


